---
title: "Reproducible Research: Peer Assessment 1"
output: github_document
---

## Read the data from /repdata-data-activity/activity.csv


```{r read_actvity_data}
activity <- read.csv("./repdata-data-activity/activity.csv", header = TRUE, colClass = c('integer', 'Date', 'integer'))
```


## What is mean total number of steps taken per day?
```{r steps_per_day}
attach(activity)
stepsbydate <- aggregate(steps ~ date, activity, na.rm=TRUE, sum)
head(stepsbydate)
```

## Barplot of the total steps taken per day

```{r barplot_steps}

barplot(stepsbydate$steps, names.arg = stepsbydate$date, ylim=c(0, 25000), 
        xlab="Date", ylab="Total Steps", col = "red", main = "Total Steps by Date")

```


## Calculate and print mean and median number of steps
```{r calculateavgact}
mean_steps <- mean(stepsbydate$steps, na.rm = TRUE)
print(mean_steps)

median_steps <- median(stepsbydate$steps, na.rm = TRUE)
print(median_steps)
```

## What is the average daily activity pattern?
```{r avg_daily_pattern}


stepsInterval <- aggregate(steps ~ interval, activity, mean)

plot(stepsInterval, type='l', main = "Average Daily Activity Pattern", col = "red",
     xlab="Interval", ylab="Steps")


```

## On average across all the days in the dataset, the 5-minute interval contains the maximum number of steps?
```{r maxsteps}
maxsteps <- stepsInterval$interval[which.max(stepsInterval$steps)]

print(maxsteps)
```

## Calculate missing values in the data set

There are many days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r missingvalues}
missingSteps <- sum(is.na(activity$steps))

print(missingSteps)
```

## Replace each missing value with the mean value of its 5-minute interval

```{r replacemissing}
activityNew <- activity
nas <- is.na(activityNew$steps)
stepInterval <- tapply(activityNew$steps, activityNew$interval, mean, na.rm=TRUE, simplify=TRUE)
activityNew$steps[nas] <- stepInterval[as.character(activityNew$interval[nas])]

sum(is.na(activityNew$steps)) ## Verify that there are no missing values
```

## Make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps (including adding the missing values from above).

```{r barplotwmissing}


allSteps <- aggregate(steps ~ date, activityNew, sum)

barplot(allSteps$steps, names.arg = allSteps$date, ylim=c(0, 25000), col = "blue", 
        xlab="Date", ylab="Total Steps", main = "Total Daily Steps Modified Dataset")

```

Mean and median values are higher after imputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

## Mean and median of the steps taken per day with new dataset

```{r avgswmissing}
mean(allSteps$steps)

median(allSteps$steps)

```



## Are there differences in activity patterns between weekdays and weekends? Using new dataset.
First figure out which are weekdays and which are weekends. 

```{r Set_Day_type}
typeofDay <- function(dates) {
  f <- function(date) {
    if (weekdays(date) %in% c("Saturday", "Sunday")) {
      "weekend"
    }
    else {
      "weekday"
    }
  }
  sapply(dates, f)
}

activityNew$typeofDay <- as.factor(typeofDay(activityNew$date))
str(activityNew)

```

##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends using activityNew data.

```{r Weekend_weekday}
library(lattice)

allstepsInterval <- aggregate(steps ~ interval + typeofDay, activityNew, mean)

xyplot(steps ~ interval | typeofDay, scales=list(cex=.8, col="purple"), data=allstepsInterval, layout=c(2,1), type='l', xlab = "Interval", ylab = "Steps", main = "Weekend Versus Weekday Steps")
```